---
title: "HW2"
author: "Yan Liu"
date: "9/15/2019"
output:
  pdf_document: default
  word_document: default
---

## 1.
### a) 
Let $X$ be an exponential r.v. with rate $1$. Using cdf inversion, write a function that generates $n$ independent samples of $X$ (we discussed this example in class, but you should do the cdf inversion yourself, not just quote our result).  Compare the speed of your sampler for $n=10^6$ with that of your language's exponential sampler (in R $\textbf{rexp}$).

### Solution:
Since $X\sim Exp(1)$, we have PDF $f(x) = e^{-x}$, $x\in [0,\infty)$, and CDF $F(x) = P(x\leq c) = \int_0^c e^{-x} dx = 1-e^{-c}$. Thus, $x = F(F^{-1}(x)) = 1-e^{-F^{-1}(x)}$. Solve for $F^{-1}(x)$, we have  $F^{-1}(x) = -ln(1-x)$
```{r}
time.start <- proc.time()
r.exp <- rexp(n=10^6)
proc.time() - time.start
```

```{r}
myExpFunc <- function(N){
  return(-log(1-runif(N)))
}
time.start <- proc.time()
my.exp <- myExpFunc(N=10^6)
proc.time() - time.start
```

### b) 
Let $X$ be a r.v. with the following pdf $f(x) = \frac{1}{2} \exp[-|x|]$,
where $\exp[-|x|] = e^{-|x|}$.  Write a function that samples from $X$ using cdf inversion.Either theoretically or through numerical examples, show that your sampler correctly samples $X$.

### Solution
$$
f(x) = \frac{1}{2}\exp(-|x|) = \left\{\begin{array}{rcl}
        \frac{1}{2}e^{-x} ,&\mbox {if}\ x\geq 0 \\
        \frac{1}{2}e^x, & \mbox{otherwise}
        \end{array}\right.\\
$$
Thus, when $c < 0$,
$$
F(x) = \Pr(x\leq c) = \int_{-\infty}^c \frac{1}{2}e^xdx = \frac{1}{2}e^c
$$
When $c\geq 0$,
$$
\begin{aligned}
F(x) &= \Pr(x\leq c) \\
&=P(x<0) + P(0\leq x\leq c) \\
&=\int_{-\infty}^0 \frac{1}{2}e^xdx\int_{0}^c \frac{1}{2}e^{-x}dx \\
&= \frac{1}{2}e^0+(-\frac{1}{2}e^{-x}|_0^c)\\
&= 1-\frac{1}{2}e^{-c}
\end{aligned}
$$
Thus,
$$
F(x) = \left\{\begin{array}{rcl}
        \frac{1}{2}e^{x} ,&\mbox {if}\ x< 0 \\
        1-\frac{1}{2}e^{-x}, & \mbox{otherwise}
        \end{array}\right.\\
$$
Let $y_1 = F(x)$ when $x<0$, $y_2 = F(x)$ when $x\geq0$.

$$
\begin{aligned}
y_1 &= \frac{1}{2}e^{x} \Longrightarrow x = log(2y_1),\ where\ y_1 < \frac{1}{2}\\
y_2 &= 1-\frac{1}{2}e^{-x} \Longrightarrow  x = -log(2-2y_2), \ where\ y_2 \geq \frac{1}{2}
\end{aligned}
$$
Thus, 
$$
F^{-1}(x) = \left\{\begin{array}{rcl}
        log(2x) ,&\mbox {if}\ x< \frac{1}{2} \\
         -log(2-2x), & \mbox{otherwise}
        \end{array}\right.\\
$$
Prove the two CDF are equivalent.
Let $C_x$ be any constant less than 0, $y$ follows the distribution determined by $F^{-1}$ above, $C_y = \frac{1}{2}e^{C_x} < \frac{1}{2}$, then,
$$
\begin{aligned}
F_y(C_y) &= \Pr(y\leq C_y) \\
&= \Pr(F_y^{-1}(y) \leq F_y^{-1}(C_y))\\
&= \Pr (x \leq log(2\times\frac{1}{2}e^{C_x} ) \\
&= \Pr (x \leq C_x) \\
&= F_x(C_x)
\end{aligned}
$$

Let $C_x$ be any constant equal to or greater than 0, $C_y = 1- \frac{1}{2}e^{-C_x} \geq \frac{1}{2}$, then,
$$
\begin{aligned}
F_y(C_y) &= \Pr(y\leq C_y) \\
&= \Pr(F_y^{-1}(y) \leq F_y^{-1}(C_y))\\
&= \Pr (x \leq -log(2- 2\times(1- \frac{1}{2}e^{-C_x}))) \\
&= \Pr (x \leq -log(e^{-C_x}))\\
&= \Pr (x \leq C_x) \\
&= F_x(C_x)
\end{aligned}
$$

## 2.
Write a function \textbf{MarkovChain(P, $s_0$, s)} that simulates a Markov chain $X(t)$ until the first time the chain is in state $s$, assuming $X(0) = s_0$.  The function should return the path of the chain from $t=0$ to when it "hits" state $s$.  You may use your language's discrete sampler (in R \textbf{sample}) or write your own.
```{r include=FALSE}
library(dplyr)
```

```{r}
#' A helper function to determine the next state using runif()
#' @param row.cumsum cumsum probabilities of current row
#' @param states list of states
#' @returns next state
roll.dice <- function(row.cumsum, states){
  # roll a dice
  dice <- runif(1)
  n <- length(states)
  # could be the initial state
  if (dice >= 0 & dice < row.cumsum[1])
      return(states[1])
  else{
       for(i in seq(1,n)){
         # could be one of the middle states
         if (i <= n-1){
                  if (dice>= row.cumsum[i] & dice < row.cumsum[i+1]){
           return(states[i+1])
                  }
         }else{
           # otherwise the last state
             return(states[n])
           }
       }
  }
}
```


```{r}
#' A function MarkovChain(P, s0, s) that simulates a Markov chain X(t) 
#' until the first time the chain is in state s, assuming X(0) = s0. 
#' @param P a transition probability matrix
#' @param s0 initial state
#' @param s target state or end state
#' @details P must satisify: a square matrix; row sum must be 1; 
#' each entry is less than or equal to 1; 
#' the row names and the column names of P are both states. 
#' The maximum steps to take is limited to 10000
#' @returns a list of states, i.e., 
#' the path of the chain from t = 0 to when it ”hits” state s.
MarkovChain <- function(P,s0,s)
{
  if (is.null(s0))
    stop("Initial state must be non-null")

  if (is.null(s))
    stop("End state must be non-null")

  if (is.null(P)) {
    stop("Transition matrix must be non-null")
  }else{
    if (nrow(P) == ncol(P) & all(P <= 1) & all(P >= 0) & rowSums(P) == 1){
      path <- list()
      states <- row.names(P)
      s.current <- which(rownames(P)==s0)
      path[1] <- s0
      row.cumsum <- cumsum(P[s.current,])
      s.next <- roll.dice(row.cumsum, states)
      numSteps <- 1
      maxSteps <- 10000
      while (s.next != s & numSteps <= maxSteps ){
        row.num.current <- which(rownames(P)==s.next)
        row.cumsum <- cumsum(P[row.num.current,])
        path[numSteps+1] <- s.next
        s.next <- roll.dice(row.cumsum, states)
        numSteps <- numSteps + 1
      }
      path[numSteps+1] <- s.next
    }
    else{
      stop("Transition matrix must be correctly specified")
    }
  }
  return (path)
}
```


```{r warning=FALSE}
states <- c("worse","bad", "okay", "good","better")
P <- matrix(c(0,.2,.3,.4,.1, 
              .5,.5,.0,0,0, 
              .05,.05,.05,.05,.8,
              .5,.5,.0,0,0,
              0,.2,.3,.4,.1), nrow = 5, ncol = 5, byrow = TRUE,
              dimnames = list(states,states))
set.seed(10)
MarkovChain(P,"worse","better")
```
```{r warning=FALSE}
set.seed(6)
MarkovChain(P,"good","better")
```

## 3.
Chutes and Ladders is a popular children's game.
Use your function from problem 2 and a Monte Carlo approach to compute $E[L]$ where $L$ is the average length, in terms of the number of die rolls, of a Chutes and Ladders game.  Use a CLT argument to determine roughly how many games you have to simulate to estimate $E[L]$ to an accuracy of $\pm 5$.
```{r}
chute.ladder <- read.csv("chutes_and_ladder_locations.csv")
game.mx <- matrix(data = rep(0,100*100), 
                  nrow = 100,
                  ncol = 100,
                  dimnames = list(seq(1,100),seq(1,100)))
for (i in 1:94){
  for (j in 1:6){
    game.mx[i, i+j] <- 1/6
  } 
}

for (i in 95:99){
  j = 1
  while (j <= 99-i){
    game.mx[i,i+j] <- 1/6
    j <- j+1
  }
  game.mx[i,100] <- 1-sum(game.mx[i,1:99])
}
```
### Take a glance
### First few entries
```{r}
game.mx[1:5,1:5]
```
### Last few entries
```{r}
game.mx[95:100,95:100]
```
### Update based on chute and ladder locations
```{r}
entrances <- chute.ladder$start
exits <- chute.ladder$end
update <- function(game.mx, entrances, exits){
    len <-  length(entrances)
    updated.mx <- game.mx
    for (i in seq(1:len)){
      updated.mx[entrances[i],] <- updated.mx[exits[i],]
    }
  return(updated.mx)
}
update.mx <- update(game.mx, entrances,exits)
```
### Take a glance
### First ladder
```{r}
update.mx[c(1,38),35:40]
```
### First chute
```{r}
update.mx[c(98,78),80:85]
```


```{r warning=FALSE}
playGame <-  function(N){
  steps = numeric()
  for (i in 1:N){
    dice.0 <- runif(1, min = 0, max = 6)
    s0 <- case_when(
                    dice.0 <= 1~38,
                    dice.0 > 1 & dice.0 <= 2~2,
                    dice.0 > 2 & dice.0 <= 3~3,
                    dice.0 > 3 & dice.0 <= 4~14,
                    dice.0 > 4 & dice.0 <= 5~5,
                    dice.0 > 5 & dice.0 <= 6~6
                    )
    
    path <- MarkovChain(update.mx, s0,100)
    steps[i] <- length(path)
  }
  return(steps)
}
set.seed(1024)
# number of times 
Ns <- seq(50, 1000, 50)
sample.means <-  numeric()
sample.sigmas <- numeric()
for (i in 1:length(Ns)){
  steps <- playGame(Ns[i])
  sample.means[i] <- mean(steps)
  sample.sigmas[i] <- sqrt(var(steps))
}

```

### Check convergence of sample mean 
```{r}
plot(Ns, sample.means, type = "b", xlab = "Number of trials")
```

### Check convergence of sample standard deviation
```{r}
plot(Ns,sample.sigmas, type = "b", xlab = "Number of trials")
```

Next, determine the number of trials that would enable the estimated $\hat E[L]$ to have an accuracy of 
$\pm 5$ compared with the true $E[L]$.Based on the central limit theorem, 
$$\lim _{n\rightarrow\infty} \sqrt n(\frac{1}{n}\sum_{i=1}^{n}\hat L_i - E(L) \sim N(0,\sigma^2)$$
Let $$\hat E[L]=\frac{1}{n}\sum_{i=1}^{n}\hat L_i$$    
In other words,    
$$\Pr(-5\leq\hat E[L] -E[L]\leq 5) = \Pr(\hat E[L]-5 \leq E[L]\leq \hat E[L]+5)$$
After some transformations, we will have:
$$Pr(\frac{-5\sqrt n}{\sigma}\leq N(0,1) \leq \frac{5\sqrt n}{\sigma})$$

Assuming the confidence level is 99%, which corresponds to a Z-score of 2.58. Let $\frac{5\sqrt n}{\sigma} = 2.58$, where $\sigma$ is estimated by $\hat \sigma$. According to the second plot above, $\hat \sigma \approx 24$. Solve the equation for $n$, we have $n\approx 153$.


